---
title: "Final Project: Get Data"
author: "Andrew Carson"
date: "December 18, 2016"
output: html_document
---
#### Get Data

Getting the data was quite difficult, as there were many different webpages and .txt or .csv files of differing formats.  Thus, automating an extraction process was quite challenging given the many caveats.  Below is the code used to extract the data from the various sources.  

Code may appear to be repeated, but upon closer inspection, you will observe slight changes to account for the slight changes in the data source.  When the code is duplicated, I had determined that simply copying the code would be faster for extracting the data.  Given the instability of the file formats, I can't even be sure that such code will run properly in the near future, and so the repeated code provides a little more flexibility to make quick changes when needed.

Note that all final .csv files are available here: (Github link)

##### Get National Summaries

In this section, I get the presidential elections data at the national level.

```{r, eval=FALSE}

library(rvest)
library(tidyr)
library(dplyr)
library(XML)

# get voting summaries for all elections
summary_df<-c()

for (i in c(1789,seq(1792,2016,4))) {
  
  #i<-2016
  url<-paste0("http://www.presidency.ucsb.edu/showelection.php?year=",i)
    
  
  #get page
  Sys.sleep(1) #crawl delay
  page<-readHTMLTable(url)
  
  if(i %in% c(1789, 1792, 1800)) {
  
    summarytable<-page[[12]]
    summarytable<-summarytable[,c(2,4,5,6)]
    summarytable_char<-data.frame(lapply(summarytable,as.character), stringsAsFactors = FALSE)
    
    final<-cbind(i,summarytable_char[,c(1,2)],NA,summarytable_char[,c(3,4)],NA,NA)
    
    
    #combined
    names(final) <- c("Year","Party", "President", "VicePresident", "ElectoralVote", "ElectoralPerc", "PopularVote", "PopularPerc")
    
    summary_df<-rbind(summary_df,final)
  } else if (i %in% c(1796)) {
    
    summarytable<-page[[12]]
    summarytable<-summarytable[c(1:2),c(2,4,5,6)]
    summarytable_char<-data.frame(lapply(summarytable,as.character), stringsAsFactors = FALSE)
    
    final<-cbind(i,summarytable_char[,c(1,2)],NA,summarytable_char[,c(3,4)],NA,NA)
    
    
    #combined
    names(final) <- c("Year","Party", "President", "VicePresident", "ElectoralVote", "ElectoralPerc", "PopularVote", "PopularPerc")
    
    summary_df<-rbind(summary_df,final)
    
  } else if(i %in% c(seq(1804,1820,4))) {
  
    summarytable<-page[[12]]
    summarytable<-summarytable[c(2,3),c(2,4,5,6,7)]
    summarytable_char<-data.frame(lapply(summarytable,as.character), stringsAsFactors = FALSE)
    
    final<-cbind(i,summarytable_char[,c(1,2,3)],summarytable_char[,c(4,5)],NA,NA)
    
    
    #combined
    names(final) <- c("Year","Party", "President", "VicePresident", "ElectoralVote", "ElectoralPerc", "PopularVote", "PopularPerc")
    
    summary_df<-rbind(summary_df,final)
  } else if(i %in% c(seq(1824,1924,4),1936)) {
  
    summarytable<-page[[12]]
    summarytable<-summarytable[c(2:nrow(summarytable)),c(2,4:9)]
    summarytable_char<-data.frame(lapply(summarytable,as.character), stringsAsFactors = FALSE)
    
    final<-cbind(i,summarytable_char)
    
    
    #combined
    names(final) <- c("Year","Party", "President", "VicePresident", "ElectoralVote", "ElectoralPerc", "PopularVote", "PopularPerc")
    
    summary_df<-rbind(summary_df,final)
  } else if(i %in% c(1928,1932,seq(1940,2016,4))) {
  
    summarytable<-page[[13]]
    summarytable<-summarytable[c(2:nrow(summarytable)),c(2,4:9)]
    summarytable_char<-data.frame(lapply(summarytable,as.character), stringsAsFactors = FALSE)
    
    final<-cbind(i,summarytable_char)
    
    
    #combined
    names(final) <- c("Year","Party", "President", "VicePresident", "ElectoralVote", "ElectoralPerc", "PopularVote", "PopularPerc")
    
    summary_df<-rbind(summary_df,final)
  }
  

print(i)
  
}


#write out summaries for later
write.csv(summary_df,"C:/Users/Andy/Desktop/Personal/Learning/CUNY/DATA607/FinalProject/summary_df.csv")

```

##### Get State Summaries

In this section, I get the presidential elections data at the state level.

```{r, eval=FALSE}


# get voting summaries for all elections
summary_df<-c()
state_header_df<-c()

for (i in seq(1824,2016,4)) {
  
  #i<-2004
  url<-paste0("http://www.presidency.ucsb.edu/showelection.php?year=",i)
    
  
  #get page
  Sys.sleep(1) #crawl delay
  page<-readHTMLTable(url)
  
 if(i %in% c(1824)) {
  
    summarytable<-page[[13]]
    summarytable_char<-data.frame(lapply(summarytable,as.character), stringsAsFactors = FALSE)
    
    #get header
    header<-cbind(i,summarytable_char[1,])
    names(header)<-c("Year","Candidate1","Candidate2","Candidate3","Candidate4")

    
    #remove headers and end rows
    summarytable_char<-summarytable_char[4:(nrow(summarytable_char)-5),]
    
    #final
    final<-cbind(i,summarytable_char)
    
    
    #combined
    names(final) <- c("Year","State", "TotalVotes", "Candidate1Votes", "Candidate1%", "Candidate1EV"
                                                  , "Candidate2Votes", "Candidate2%", "Candidate2EV"
                                                  , "Candidate3Votes", "Candidate3%", "Candidate3EV"
                                                  , "Candidate4Votes", "Candidate4%", "Candidate4EV"
                      )
    
    summary_df<-rbind(summary_df,final)
    state_header_df<-rbind(state_header_df,header)
    

    
    
  } else if(i %in% c(seq(1828,1924,4),1936)) {
  
    summarytable<-page[[13]]
    summarytable_char<-data.frame(lapply(summarytable,as.character), stringsAsFactors = FALSE)
    
    #get header
    header<-cbind(i,summarytable_char[1,])
    header<-cbind(header,matrix(nrow=1,ncol= (15 - ncol(header))))
    names(header)<-c("Year","Candidate1","Candidate2","Candidate3","Candidate4")
    

    
    #remove headers and end rows
    summarytable_char<-summarytable_char[4:(nrow(summarytable_char)-2),]
    
    #final
    final<-cbind(i,summarytable_char)
    final<-cbind(final,matrix(nrow = nrow(final), ncol = (15 - ncol(final))))
    
    #combined
    names(final) <- c("Year","State", "TotalVotes", "Candidate1Votes", "Candidate1%", "Candidate1EV"
                                                  , "Candidate2Votes", "Candidate2%", "Candidate2EV"
                                                  , "Candidate3Votes", "Candidate3%", "Candidate3EV"
                                                  , "Candidate4Votes", "Candidate4%", "Candidate4EV"
                      )
    
    summary_df<-rbind(summary_df,final)
    state_header_df<-rbind(state_header_df,header)
    
    
  } else if(i %in% c(1928,1932,1940,1944,1952,1956)) {
  
    summarytable<-page[[12]]
    summarytable_char<-data.frame(lapply(summarytable,as.character), stringsAsFactors = FALSE)
    
    
    #get header
    summarytable_char<-summarytable_char[7:nrow(summarytable_char),]
    header<-cbind(i,summarytable_char[1,])
    header<-cbind(header,matrix(nrow=1,ncol= (15 - ncol(header))))
    names(header)<-c("Year","Candidate1","Candidate2","Candidate3","Candidate4")
    

    
    #remove headers and end rows
    summarytable_char<-summarytable_char[5:(nrow(summarytable_char)-2),]
    
    #final
    final<-cbind(i,summarytable_char)
    final<-cbind(final,matrix(nrow = nrow(final), ncol = (15 - ncol(final))))
    
    #combined
    names(final) <- c("Year","State", "TotalVotes", "Candidate1Votes", "Candidate1%", "Candidate1EV"
                                                  , "Candidate2Votes", "Candidate2%", "Candidate2EV"
                                                  , "Candidate3Votes", "Candidate3%", "Candidate3EV"
                                                  , "Candidate4Votes", "Candidate4%", "Candidate4EV"
                      )
    
    summary_df<-rbind(summary_df,final)
    state_header_df<-rbind(state_header_df,header)
    
    
  } else if(i %in% c(1948)) {
  
    summarytable<-page[[12]]
    summarytable_char<-data.frame(lapply(summarytable,as.character), stringsAsFactors = FALSE)
    
    
    #get header
    summarytable_char<-summarytable_char[8:nrow(summarytable_char),]
    header<-cbind(i,summarytable_char[1,])
    header<-cbind(header,matrix(nrow=1,ncol= (15 - ncol(header))))
    names(header)<-c("Year","Candidate1","Candidate2","Candidate3","Candidate4")
    

    
    #remove headers and end rows
    summarytable_char<-summarytable_char[5:(nrow(summarytable_char)-2),]
    
    #final
    final<-cbind(i,summarytable_char)
    final<-cbind(final,matrix(nrow = nrow(final), ncol = (15 - ncol(final))))
    
    #combined
    names(final) <- c("Year","State", "TotalVotes", "Candidate1Votes", "Candidate1%", "Candidate1EV"
                                                  , "Candidate2Votes", "Candidate2%", "Candidate2EV"
                                                  , "Candidate3Votes", "Candidate3%", "Candidate3EV"
                                                  , "Candidate4Votes", "Candidate4%", "Candidate4EV"
                      )
    
    summary_df<-rbind(summary_df,final)
    state_header_df<-rbind(state_header_df,header)
    
    
  } else if(i %in% c(1960, 1968,1980,1992,1996,2000)) {
  
    summarytable<-page[[12]]
    summarytable_char<-data.frame(lapply(summarytable,as.character), stringsAsFactors = FALSE)
    
    
    #get header
    summarytable_char<-summarytable_char[6:nrow(summarytable_char),]
    header<-cbind(i,summarytable_char[1,])
    header$V1<-NULL
    
    header<-cbind(header,matrix(nrow=1,ncol= (15 - ncol(header))))
    names(header)<-c("Year","Candidate1","Candidate2","Candidate3","Candidate4")
    

    
    #remove headers and end rows
    summarytable_char<-summarytable_char[4:(nrow(summarytable_char)-2),]
    
    #final
    final<-cbind(i,summarytable_char)
    final<-cbind(final,matrix(nrow = nrow(final), ncol = (15 - ncol(final))))
    
    #combined
    names(final) <- c("Year","State", "TotalVotes", "Candidate1Votes", "Candidate1%", "Candidate1EV"
                                                  , "Candidate2Votes", "Candidate2%", "Candidate2EV"
                                                  , "Candidate3Votes", "Candidate3%", "Candidate3EV"
                                                  , "Candidate4Votes", "Candidate4%", "Candidate4EV"
                      )
    
    summary_df<-rbind(summary_df,final)
    state_header_df<-rbind(state_header_df,header)
    
    
  } else if(i %in% c(1964,1972,1976,1984,1988)) {
  
    summarytable<-page[[12]]
    summarytable_char<-data.frame(lapply(summarytable,as.character), stringsAsFactors = FALSE)
    
    
    #get header
    summarytable_char<-summarytable_char[6:nrow(summarytable_char),]
    header<-cbind(i,summarytable_char[1,])
    header<-cbind(header,matrix(nrow=1,ncol= (15 - ncol(header))))
    names(header)<-c("Year","Candidate1","Candidate2","Candidate3","Candidate4")
    

    
    #remove headers and end rows
    summarytable_char<-summarytable_char[5:(nrow(summarytable_char)-2),]
    
    #final
    final<-cbind(i,summarytable_char)
    final<-cbind(final,matrix(nrow = nrow(final), ncol = (15 - ncol(final))))
    
    #combined
    names(final) <- c("Year","State", "TotalVotes", "Candidate1Votes", "Candidate1%", "Candidate1EV"
                                                  , "Candidate2Votes", "Candidate2%", "Candidate2EV"
                                                  , "Candidate3Votes", "Candidate3%", "Candidate3EV"
                                                  , "Candidate4Votes", "Candidate4%", "Candidate4EV"
                      )
    
    summary_df<-rbind(summary_df,final)
    state_header_df<-rbind(state_header_df,header)
    
    
  } else if(i %in% c(2004,2008)) {
  
    summarytable<-page[[12]]
    summarytable_char<-data.frame(lapply(summarytable,as.character), stringsAsFactors = FALSE)
    
    
    #get header
    summarytable_char<-summarytable_char[6:nrow(summarytable_char),]
    header<-cbind(i,summarytable_char[1,])
    header<-cbind(header,matrix(nrow=1,ncol= (15 - ncol(header))))
    names(header)<-c("Year","Candidate1","Candidate2","Candidate3","Candidate4")
    

    
    #remove headers and end rows
    summarytable_char<-summarytable_char[4:(nrow(summarytable_char)-2),]
    
    #final
    final<-cbind(i,summarytable_char)
    final<-cbind(final,matrix(nrow = nrow(final), ncol = (15 - ncol(final))))
    
    #combined
    names(final) <- c("Year","State", "TotalVotes", "Candidate1Votes", "Candidate1%", "Candidate1EV"
                                                  , "Candidate2Votes", "Candidate2%", "Candidate2EV"
                                                  , "Candidate3Votes", "Candidate3%", "Candidate3EV"
                                                  , "Candidate4Votes", "Candidate4%", "Candidate4EV"
                      )
    
    summary_df<-rbind(summary_df,final)
    state_header_df<-rbind(state_header_df,header)
    
    
  } else if(i %in% c(2012)) {
  
    summarytable<-page[[12]]
    summarytable_char<-data.frame(lapply(summarytable,as.character), stringsAsFactors = FALSE)
    
    
    #get header
    summarytable_char<-summarytable_char[8:nrow(summarytable_char),]
    header<-cbind(i,summarytable_char[1,])
    header$V1<-NULL
    header<-cbind(header,matrix(nrow=1,ncol= (15 - ncol(header))))
    names(header)<-c("Year","Candidate1","Candidate2","Candidate3","Candidate4")
    

    
    #remove headers and end rows
    summarytable_char<-summarytable_char[4:(nrow(summarytable_char)-2),]
    
    #final
    final<-cbind(i,summarytable_char)
    final<-cbind(final,matrix(nrow = nrow(final), ncol = (15 - ncol(final))))
    
    #combined
    names(final) <- c("Year","State", "TotalVotes", "Candidate1Votes", "Candidate1%", "Candidate1EV"
                                                  , "Candidate2Votes", "Candidate2%", "Candidate2EV"
                                                  , "Candidate3Votes", "Candidate3%", "Candidate3EV"
                                                  , "Candidate4Votes", "Candidate4%", "Candidate4EV"
                      )
    
    summary_df<-rbind(summary_df,final)
    state_header_df<-rbind(state_header_df,header)
    
    
  } else if(i %in% c(2016)) {
  
    summarytable<-page[[12]]
    summarytable_char<-data.frame(lapply(summarytable,as.character), stringsAsFactors = FALSE)
    
    
    #get header
    summarytable_char<-summarytable_char[11:nrow(summarytable_char),]
    header<-cbind(i,summarytable_char[1,])
    header<-cbind(header,matrix(nrow=1,ncol= (15 - ncol(header))))
    names(header)<-c("Year","Candidate1","Candidate2","Candidate3","Candidate4")
    

    
    #remove headers and end rows
    summarytable_char<-summarytable_char[4:(nrow(summarytable_char)-2),]
    
    #final
    final<-cbind(i,summarytable_char)
    final<-cbind(final,matrix(nrow = nrow(final), ncol = (15 - ncol(final))))
    
    #combined
    names(final) <- c("Year","State", "TotalVotes", "Candidate1Votes", "Candidate1%", "Candidate1EV"
                                                  , "Candidate2Votes", "Candidate2%", "Candidate2EV"
                                                  , "Candidate3Votes", "Candidate3%", "Candidate3EV"
                                                  , "Candidate4Votes", "Candidate4%", "Candidate4EV"
                      )
    
    summary_df<-rbind(summary_df,final)
    state_header_df<-rbind(state_header_df,header)
    
    
  }

print(i)
  
}


#write out summaries for later
write.csv(summary_df,"C:/Users/Andy/Desktop/Personal/Learning/CUNY/DATA607/FinalProject/summary_state_df.csv")

write.csv(state_header_df,"C:/Users/Andy/Desktop/Personal/Learning/CUNY/DATA607/FinalProject/state_header_df.csv")

```


  
##### Get National Population Data: 1900 - 2016

In this section, I get the yearly population data at the national level.
  
```{r, eval = FALSE}

library(stringr)

######### national population 1900-1999
national_population<-readLines("http://www.census.gov/popest/data/national/totals/pre-1980/tables/popclockest.txt")

#clean rows
national_population<-national_population[11:119]
national_population<-national_population[which(national_population != "")]
nat_pop_1_85<-str_split(national_population[1:85],"     ", simplify = TRUE)
nat_pop_86_100<-str_split(national_population[86:100],"       ", simplify = TRUE)
nat_pop_1_85<-nat_pop_1_85[,c(1,2,4,6)]
nat_pop_86_100<-nat_pop_86_100[,c(1,2,3,5)]

#transform into dataframe
national_population_clean<- rbind(nat_pop_1_85,nat_pop_86_100)
national_population_clean<-data.frame(apply(national_population_clean,c(1, 2),str_trim), stringsAsFactors = FALSE)
names(national_population_clean)<-c("Date","NationalPopulation","PopChange","AvgAnnualPercChange")

#write out for later use
write.csv(national_population_clean,"C:/Users/Andy/Desktop/Personal/Learning/CUNY/DATA607/FinalProject/national_population_1900_1999.csv")


######### national population 2000 - 2010
national_population<-read.csv("http://www.census.gov/popest/data/intercensal/national/files/US-EST00INT-TOT.csv")

#write out for later use
write.csv(national_population,"C:/Users/Andy/Desktop/Personal/Learning/CUNY/DATA607/FinalProject/national_population_2000_2010.csv")


######### national population 2010 - 2016
#sourced from: http://factfinder.census.gov/bkmk/table/1.0/en/PEP/2015/PEPMONTHN/

national_population<-read.csv("C:/Users/Andy/Desktop/Personal/Learning/CUNY/DATA607/FinalProject/national_population_2010_2016.csv")

######## combine into single source file for 1900 - 2016
#1900-1999
nat_pop_1900_1999<-read.csv("C:/Users/Andy/Desktop/Personal/Learning/CUNY/DATA607/FinalProject/national_population_1900_1999.csv")
nat_pop_1900_1999$X<-NULL
nat_pop_1900_1999<-nat_pop_1900_1999[,c(1,2)]

#2000-2010
nat_pop_2000_2010<-read.csv("C:/Users/Andy/Desktop/Personal/Learning/CUNY/DATA607/FinalProject/national_population_2000_2010.csv")
nat_pop_2000_2010$X<-NULL
nat_pop_2000_2010<-nat_pop_2000_2010[which(nat_pop_2000_2010$MONTH==7),]
nat_pop_2000_2010$Date<-paste0("July 1, ",nat_pop_2000_2010$YEAR)
nat_pop_2000_2010<-nat_pop_2000_2010[,c(4,3)]
names(nat_pop_2000_2010)<-c("Date","NationalPopulation")

#2010-2016
nat_pop_2010_2016<-read.csv("C:/Users/Andy/Desktop/Personal/Learning/CUNY/DATA607/FinalProject/national_population_2010_2016.csv")
nat_pop_2010_2016<-nat_pop_2010_2016[which(substring(nat_pop_2010_2016$"Year.and.Month",1,1)==7),]
nat_pop_2010_2016$Date<-paste0("July 1, ",substring(nat_pop_2010_2016$"Year.and.Month",5,8))
nat_pop_2010_2016<-nat_pop_2010_2016[2:nrow(nat_pop_2010_2016),c(7,2)]
names(nat_pop_2010_2016)<-c("Date","NationalPopulation")


#combine
national_population_1900_2016<-
  rbind(data.frame(lapply(nat_pop_1900_1999,as.character), stringsAsFactors = FALSE)
        ,data.frame(lapply(nat_pop_2000_2010,as.character), stringsAsFactors = FALSE)
        ,data.frame(lapply(nat_pop_2010_2016,as.character), stringsAsFactors = FALSE))


#write out for later use
write.csv(national_population_1900_2016,"C:/Users/Andy/Desktop/Personal/Learning/CUNY/DATA607/FinalProject/national_population_1900_2016.csv")

```



##### Get State Population Data: 1900 - 2016

In this section, I get the yearly population data at the state level.

```{r, eval=FALSE}
#1900 - 1990

############# 1900 - 1909 : data in thousands, July 1st estimates
data<-readLines("http://www.census.gov/popest/data/state/asrh/1980s/tables/st0009ts.txt")

data1<-data[18:72]
data2<-data[76:130]

data1<-str_split(data1,"[ ]{2,}", simplify = TRUE)
data2<-str_split(data2,"[ ]{2,}", simplify = TRUE)

data_combined<-cbind(data1,data2[,2:5])
data_combined<-data.frame(data_combined,stringsAsFactors = FALSE)
names(data_combined)<-c("State",seq(1900,1909,1))
 
#write out
write.csv(data_combined,"C:/Users/Andy/Desktop/Personal/Learning/CUNY/DATA607/FinalProject/state_population_1900_1909.csv")


############## 1910 - 1919 : data in thousands, July 1st estimates
data<-readLines("http://www.census.gov/popest/data/state/asrh/1980s/tables/st1019ts.txt")

data1<-data[18:72]
data2<-data[76:130]

data1<-str_split(data1,"[ ]{2,}", simplify = TRUE)
data2<-str_split(data2,"[ ]{2,}", simplify = TRUE)

data_combined<-cbind(data1,data2[,2:5])
data_combined<-data.frame(data_combined,stringsAsFactors = FALSE)
names(data_combined)<-c("State",seq(1910,1919,1))
 
#write out
write.csv(data_combined,"C:/Users/Andy/Desktop/Personal/Learning/CUNY/DATA607/FinalProject/state_population_1910_1919.csv")

############## 1920 - 1929 : data in thousands, July 1st estimates
data<-readLines("http://www.census.gov/popest/data/state/asrh/1980s/tables/st2029ts.txt")

data1<-data[18:72]
data2<-data[76:130]

data1<-str_split(data1,"[ ]{2,}", simplify = TRUE)
data2<-str_split(data2,"[ ]{2,}", simplify = TRUE)

data_combined<-cbind(data1,data2[,2:5])
data_combined<-data.frame(data_combined,stringsAsFactors = FALSE)
names(data_combined)<-c("State",seq(1920,1929,1))
 
#write out
write.csv(data_combined,"C:/Users/Andy/Desktop/Personal/Learning/CUNY/DATA607/FinalProject/state_population_1920_1929.csv")

############## 1930 - 1939 : data in thousands, July 1st estimates
data<-readLines("http://www.census.gov/popest/data/state/asrh/1980s/tables/st3039ts.txt")

data1<-data[18:72]
data2<-data[77:131]

data1<-str_split(data1,"[ ]{2,}", simplify = TRUE)
data2<-str_split(data2,"[ ]{2,}", simplify = TRUE)

data_combined<-cbind(data1,data2[,2:5])
data_combined<-data.frame(data_combined,stringsAsFactors = FALSE)
names(data_combined)<-c("State",seq(1930,1939,1))
 
#write out
write.csv(data_combined,"C:/Users/Andy/Desktop/Personal/Learning/CUNY/DATA607/FinalProject/state_population_1930_1939.csv")

############## 1940 - 1949 : data in thousands, July 1st estimates
data<-readLines("http://www.census.gov/popest/data/state/asrh/1980s/tables/st4049ts.txt")

data1<-data[16:70]
data2<-data[74:128]

data1<-str_split(data1,"[ ]{2,}", simplify = TRUE)
data2<-str_split(data2,"[ ]{2,}", simplify = TRUE)

data_combined<-cbind(data1,data2[,2:5])
data_combined<-data.frame(data_combined,stringsAsFactors = FALSE)
names(data_combined)<-c("State",seq(1940,1949,1))
 
#write out
write.csv(data_combined,"C:/Users/Andy/Desktop/Personal/Learning/CUNY/DATA607/FinalProject/state_population_1940_1949.csv")

############## 1950 - 1959 : data in thousands, July 1st estimates
data<-readLines("http://www.census.gov/popest/data/state/asrh/1980s/tables/st5060ts.txt")

data1<-data[19:78]
data2<-data[84:143]

data1<-str_split(data1,"[ ]{2,}", simplify = TRUE)
data2<-str_split(data2,"[ ]{2,}", simplify = TRUE)

data_combined<-cbind(data1,data2[,2:7])
data_combined<-data.frame(data_combined,stringsAsFactors = FALSE)
names(data_combined)<-c("State","1950census",seq(1950,1959,1),"1960census")
 
#write out
write.csv(data_combined,"C:/Users/Andy/Desktop/Personal/Learning/CUNY/DATA607/FinalProject/state_population_1950_1959.csv")

############## 1960 - 1969 : data in thousands, July 1st estimates
data<-readLines("http://www.census.gov/popest/data/state/asrh/1980s/tables/st6070ts.txt")

data1<-data[19:75]
data2<-data[81:137]

data1<-str_split(data1,"[ ]{2,}", simplify = TRUE)
data2<-str_split(data2,"[ ]{2,}", simplify = TRUE)

data_combined<-cbind(data1,data2[,2:7])
data_combined<-data.frame(data_combined,stringsAsFactors = FALSE)
names(data_combined)<-c("State","1960census",seq(1960,1969,1),"1970census")
 
#write out
write.csv(data_combined,"C:/Users/Andy/Desktop/Personal/Learning/CUNY/DATA607/FinalProject/state_population_1960_1969.csv")

############## 1970 - 1979 : data in thousands, July 1st estimates
data<-readLines("http://www.census.gov/popest/data/state/asrh/1980s/tables/st7080ts.txt")

data1<-data[15:66]
data2<-data[68:119]

data1<-str_split(data1,"[ ]{1,}", simplify = TRUE)
data2<-str_split(data2,"[ ]{1,}", simplify = TRUE)

data1[52,]<-data1[52,c(9,1:8)]
data2[52,]<-data2[52,c(8,1:7)]

data_combined<-cbind(data1[,3:9],data2[,4:8])
data_combined<-data.frame(data_combined,stringsAsFactors = FALSE)
names(data_combined)<-c("State",seq(1970,1980,1))
 
#write out
write.csv(data_combined,"C:/Users/Andy/Desktop/Personal/Learning/CUNY/DATA607/FinalProject/state_population_1970_1979.csv")

############## 1980 - 1989 : data in thousands, July 1st estimates
data<-readLines("http://www.census.gov/popest/data/state/asrh/1980s/tables/st8090ts.txt")

data1<-data[11:62]
data2<-data[70:121]

data1<-str_split(data1,"[ ]{1,}", simplify = TRUE)
data2<-str_split(data2,"[ ]{1,}", simplify = TRUE)


data_combined<-cbind(data1,data2[,2:7])
data_combined<-data.frame(data_combined,stringsAsFactors = FALSE)
names(data_combined)<-c("State",seq(1980,1990,1))
 
#write out
write.csv(data_combined,"C:/Users/Andy/Desktop/Personal/Learning/CUNY/DATA607/FinalProject/state_population_1980_1989.csv")
  

#################################

# 1990 - 1999
data<-readLines("http://www.census.gov/popest/data/state/totals/1990s/tables/ST-99-03.txt")

data1<-data[14:78]
data2<-data[88:152]

data1<-str_split(data1,"[ ]{2,}", simplify = TRUE)
data2<-str_split(data2,"[ ]{2,}", simplify = TRUE)

data_combined<-cbind(data1[,3:9],data2[,4:8])
data_combined<-data.frame(data_combined,stringsAsFactors = FALSE)
names(data_combined)<-c("State",seq(1999,1990,-1),"census1990")

#write out for using later
write.csv(data_combined,"C:/Users/Andy/Desktop/Personal/Learning/CUNY/DATA607/FinalProject/state_population_1990_1999.csv")


# 2000 - 2009
data<-read.csv("http://www.census.gov/popest/data/intercensal/state/tables/ST-EST00INT-01.csv")
data<-data[4:59,]
names(data)<-c("State","census2000",seq(2000,2009,1),"census2010",2010)

#write out for later use
write.csv(data,"C:/Users/Andy/Desktop/Personal/Learning/CUNY/DATA607/FinalProject/state_population_2000_2009.csv")


#2010 - 2016
#sourced from : http://factfinder.census.gov/bkmk/table/1.0/en/PEP/2015/PEPANNRES/0100000US|0100000US.04000|0200000US1|0200000US2|0200000US3|0200000US4
data<-read.csv("C:/Users/Andy/Desktop/Personal/Learning/CUNY/DATA607/FinalProject/state_population_2010_2015_source.csv")
names(data)<-c("State","census2010","EstimatesBase",seq(2010,2015,1))
data<-data[2:57,]

#write out for later use
write.csv(data,"C:/Users/Andy/Desktop/Personal/Learning/CUNY/DATA607/FinalProject/state_population_2010_2015.csv")


```

#### Conclusion

This was probably the most difficult part of the project for me, due to the time consuming nature of gathering the data and the intricate level of detail needed to correctly extract it.  There is an often quoted assertion that a data scientist spends 80% of his or her time just getting, cleaning, and transforming the data.  In this case, this was certainly true.

