---
title: "DATA607 Project 1: Chess Tournament"
author: "Andrew Carson"
date: "September 13, 2016"
output: html_document
---


### Project Description
In this project, you're given a text file with chess tournament results where the information has some structure. Your job is to create an R Markdown file that generates a .CSV file (that could for example be imported into a SQL database) with the following information for all of the players:

    Player's Name, Player's State, Total Number of Points, Player's Pre-Rating, and Average Pre Chess Rating of Opponents

For the first player, the information would be:
Gary Hua, ON, 6.0, 1794, 1605

1605 was calculated by using the pre-tournament opponents' ratings of 1436, 1563, 1600, 1610, 1649, 1663, 1716, and dividing by the total number of games played.

All of your code should be in an R markdown file (and published to rpubs.com), with your data accessible for the person running the script.

###Solution
```{r code, eval=TRUE, echo=TRUE}
library(stringr)

#read data
raw_data<-read.delim("https://raw.githubusercontent.com/anrcarson/CUNY-MSDA/master/DATA607/Project1/tournamentinfo.txt",header = FALSE, stringsAsFactors = FALSE)

#view data
#raw_data

#remove header
data_no_header<-raw_data[5:196,]

#get name row
data_name_rows<-data_no_header[seq(from=1,to=192, by=3)]
name_df<-
 as.data.frame(matrix(unlist(str_split(data_name_rows,"\\|")),nrow=64, byrow=T), stringsAsFactors = FALSE) #credit: http://ggorjan.blogspot.com/2011/01/converting-strsplit-output-to-dataframe.html

#set names
names(name_df)<-c("PairNum","PlayerName","TotalPoints","Round1","Round2","Round3","Round4","Round5","Round6","Round7","Empty")
name_df$Empty<-NULL

#get state row
data_state_rows<-data_no_header[seq(from=2,to=192, by=3)]
state_df<-
 as.data.frame(matrix(unlist(str_split(data_state_rows,"\\|")),nrow=64, byrow=T), stringsAsFactors = FALSE) #credit: http://ggorjan.blogspot.com/2011/01/converting-strsplit-output-to-dataframe.html

names(state_df)<-c("PlayerState","IdRating","TotalPoints_2","Round1_2","Round2_2","Round3_2","Round4_2","Round5_2","Round6_2","Round7_2","Empty")
state_df$Empty<-NULL


#combine the two dfs
df<-cbind(name_df,state_df)

#get pre-rating
df$PreRating<-matrix(unlist(str_extract_all(df$IdRating,"[:digit:]{3,}")), nrow=64, byrow=T)[,2]
df$PreRating<-as.numeric(df$PreRating)

#get opponents
df$Round1_Opp<-str_extract(df$Round1,"[:digit:]{1,}")
df$Round2_Opp<-str_extract(df$Round2,"[:digit:]{1,}")
df$Round3_Opp<-str_extract(df$Round3,"[:digit:]{1,}")
df$Round4_Opp<-str_extract(df$Round4,"[:digit:]{1,}")
df$Round5_Opp<-str_extract(df$Round5,"[:digit:]{1,}")
df$Round6_Opp<-str_extract(df$Round6,"[:digit:]{1,}")
df$Round7_Opp<-str_extract(df$Round7,"[:digit:]{1,}")


#trim pairnum
df$PairNum<-str_trim(df$PairNum)

#get average pre-rating of opponents
average_prerating <- c()
for(i in 1:nrow(df)) {
   #i<-44
  average_prerating<-c(average_prerating,
    round(                   
    mean(
      c(
     df$PreRating[which(df$PairNum == df$Round1_Opp[i])]
     ,df$PreRating[which(df$PairNum == df$Round2_Opp[i])]
     ,df$PreRating[which(df$PairNum == df$Round3_Opp[i])]
     ,df$PreRating[which(df$PairNum == df$Round4_Opp[i])]
     ,df$PreRating[which(df$PairNum == df$Round5_Opp[i])]
     ,df$PreRating[which(df$PairNum == df$Round6_Opp[i])]
     ,df$PreRating[which(df$PairNum == df$Round7_Opp[i])]
      )
  ,na.rm = TRUE, trim=0)
  ))
  
}

df$AveragePreRating_Opponents<-average_prerating


#assemble final df
final_df<-data.frame(
              cbind(
              str_trim(df$PlayerName)
              ,str_trim(df$PlayerState)
              ,str_trim(df$TotalPoints)
              ,str_trim(df$PreRating)
              ,str_trim(df$AveragePreRating_Opponents)
                ) #cbind
                ,stringsAsFactors = FALSE)

names(final_df)<-c("PlayerName","PlayerState","PlayerTotalPoints","PlayerPreRating","AvgPreRating_Opponents")
```

```{r write, eval=FALSE}
#write the csv file
write.csv(final_df, file="tournamentinfo_clean.csv", row.names = FALSE)

```


### Add Extra Columns
```{r extra}

#get post-rating
final_df$PlayerPostRating<-
  matrix(unlist(str_extract_all(df$IdRating,"[:digit:]{3,}")), nrow=64, byrow=T)[,3]
final_df$PlayerPostRating<-as.numeric(final_df$PlayerPostRating)

#make numeric
final_df$PlayerPreRating<-as.numeric(final_df$PlayerPreRating)
final_df$AvgPreRating_Opponents<-as.numeric(final_df$AvgPreRating_Opponents)

#get PostRating - PreRating difference
final_df$PlayerPostPreDiff<-final_df$PlayerPostRating - final_df$PlayerPreRating

#get PlayerPrerating - AvgPreRating_Opponents diff
final_df$PlayerPreRateAvgOppDiff<- final_df$PlayerPreRating - final_df$AvgPreRating_Opponents

```



### Data Exploration

Here is the data.
```{r explore0}
library(DT)
datatable(final_df)

```

Pre-rating is skewed left.  The mean is 1378 with a low of 377 and a max of 1764.  The standard deviation is 265.
``` {R explore1}
hist(final_df$PlayerPreRating, xlim = c(0,2000), ylim = c(0,10), breaks = seq(0,2000,50))
summary(final_df$PlayerPreRating)
sd(final_df$PlayerPreRating)
```

Post-rating is more normal than pre-rating, though still skewed left.  The mean is 1398 with a min of 878 and a max of 1817.  The standard deviation is 231.
``` {R explore2}
hist(final_df$PlayerPostRating,  xlim = c(0,2000), ylim = c(0,10), breaks = seq(0,2000,50))
summary(final_df$PlayerPostRating)
sd(final_df$PlayerPostRating)
```

Looking at a boxplot for each, the distribution doesn't change much, except that PostRating no longer has the outlier that PreRating had.
```{r explore3}
boxplot(final_df$PlayerPreRating, final_df$PlayerPostRating, names = c("PreRating","PostRating"))
```

Viewing the difference between the pre-rating and the post-rating, we see that it is not normall distributed.  It is highly skewed right.  The min is -94, the max is 699, and the median is 0 while the mean is 19.67.  Interestingly, the same number of players gained in rating (31) as did lose in rating (31), while 2 players had no change.

```{r explore4}
hist(final_df$PlayerPostPreDiff)
summary(final_df$PlayerPostPreDiff)
sum(final_df$PlayerPostPreDiff < 0)/length(final_df$PlayerPostPreDiff)
sum(final_df$PlayerPostPreDiff > 0)/length(final_df$PlayerPostPreDiff)
sum(final_df$PlayerPostPreDiff == 0)/length(final_df$PlayerPostPreDiff)
```

